Distance Sampling in \proglang{R}

### 2016-03-31 paper.tex
remove temp files add to .gitignore
stop ignoring the paper.tex file now

\author{
David L Miller\\University of St Andrews \And Eric Rexstad\\University of St Andrews \And Jeffrey L Laake\\National Marine Mammal Laboratory
}
\title{Distance Sampling in \proglang{R}}
\Keywords{distance sampling, abundance estimation, line transects, point transects, \proglang{R}}

\Abstract{
The abstract of the article.
}

\Plainauthor{David L Miller, Eric Rexstad, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R}
\Plainkeywords{distance sampling, abundance estimation, line transects, point transects, R}

@@

\Address{
    David L Miller\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:dave@ninepointeightone.net}{\nolinkurl{dave@ninepointeightone.net}}\\
  URL: \url{http://converged.yt}\\~\\
      Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:Eric.Rexstad@st-andrews.ac.uk}{\nolinkurl{Eric.Rexstad@st-andrews.ac.uk}}\\

      Jeffrey L Laake\\
  National Marine Mammal Laboratory\\
  Alaska Fisheries Science Center 7600 Sand Point Way N.E., Seattle, WA
  98115, USA\\
  E-mail: \href{mailto:Jeff.Laake@noaa.gov}{\nolinkurl{Jeff.Laake@noaa.gov}}\\

@@

\usepackage{amsmath} \usepackage{amssymb} \usepackage{longtable}
\usepackage{booktabs}

@@

Distance sampling (Buckland et al. 2001; Buckland et al. 2004)
encompasses a suite of field methods and statistical models used to
estimate the abundance of biological populations. Distance sampling
field procedure can be thought of as an extension of plot sampling,
where we wish to take into account the decreasing probability of
detecting objects at increasing distance from the sampler. We do this by
building a model for detectability and use quantities calculated from
the detection function to adjust the observed counts to obtain an
estimate of abundance.

For many years distance sampling analyses have been available via the
Windows program Distance (or ``DISTANCE''); for clarity henceforth
``Distance for Windows'' Thomas et al. (2010)). From version 5 of
Distance for Windows, R packages have been included to perform
particular analyses (CITE Distance user manual). This paper shows how to
fit detection functions, perform model checking and selection and
estimate abundance in \proglang{R} using the package \pkg{Distance}.
\pkg{Distance} is a wrapper package around the more complex (and more
powerful) \pkg{mrds} and offers a subset of the analyses possible with
that package.

@@

Census-type surveys (quadrat or strip transects assuming perfect
detectability within strips or quadrats) are inefficient (requiring
considerable field effort) and we should expect that not all objects
(animals, plants, dung, etc) can be observed. Accounting for imperfect
detectability is an important consideration when obtaining accurate
estimates of abundance (Lahoz-Monfort, Guillera-Arroita, and Wintle
2013). Using the extra information gained by recording distance from the
sampler to the observation, it is possible to model detectability.
Because we expect detectability to decrease with increasing distance
from the sampler, we model detectability as a function of distance (plus
perhaps other covariates, see below). We use the probabilities of
detection calculated from the detection function to estimate an
abundance for the area covered by the survey, which we then scale-up to
the study region of interest.

Distance sampling surveys are conducted using two transect types: line
and point transects. In line transect sampling observers walk (or fly,
sail, etc) lines observing objects and recording the distances to the
line; whereas in point transect sampling observers remain stationary at
a location and record distances from that point. Field methods are
chosen to be suitable to species and habitat constraints (Buckland et
al. 2015).

@@

@@

\caption{Left side plots show an example of a survey of an area containing a population of 500 objects, blue indicates sampler placement (top lines, bottom points) and red dots indicate detected individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\label{fig:pointslines}}\label{fig:points-and-lines}
\end{figure}
\end{CodeChunk}

For both points and lines, given the geometry of the sampler (see
``Detection functions''), the histogram of distances should show a
decreasing number of observations with increasing distance from the
sampler. For line transects we expect objects to be distributed
uniformly with respect to distance from the line with detectability
causing the number of detections to decrease. For point transects, we
recognise that as distance from the point increases, the area of the
circle encompassed increases with distance squared; hence the number of
objects available to be detected is a linearly increasing function of
distance from the point.

Using this histogram we can crudely estimate the drop-off in
detectability by eye by tracing a line that approximates the tops of the
histogram bars -- this is the detection function. \pkg{Distance}
estimates the parameters for a fixed-form detection function using
maximum likelihood estimation. We address possible models below.

Figure \ref{fig:pointslines} shows sampling of a population of 500
individuals using line and point transects (left column) and their
corresponding histograms (right column).

\subsection{Data}\label{data}

We demonstrate \pkg{Distance} using two data sets: one line transect and
one point transect. These data sets have been chosen to be
representative of data seen in practice.

@@

The line transect data is based on a survey of Antarctic minke whales
(\emph{Balaenoptera bonaerensis}). The data are simulated from models
fitted to data from the International Whaling Commission's International
Decade of Cetacean Research Southern Ocean Whale and Ecosystem Research
(IWC IDCR-SOWER) programme 1992-1993 austral summer surveys (Branch and
Butterworth 2001). They consist of 99 observations and include
geographical strata (near or distant from ice edge) and transect lengths
on the effort expended.

@@

The point transect data set consists of 1485 observations of Amakihi
(\emph{Hemignathus virens}; a Hawaiian songbird), collected at 41 points
between 1992 and 1995. The data include distances and two covariates
collected during the survey: observer (a three level factor), time after
sunrise (transformed to minutes (continuous) or hours (factor)
covariates). Data are analysed comprehensively in Marques et al. (2007).

\subsection{The rest of the paper}\label{the-rest-of-the-paper}

The rest of the paper has this structure: we describe data organisation
for use with \pkg{Distance}; models for the detection function are
described in terms of formulation and examples of fitting in
\proglang{R}. We then examine model checking, goodness of fit and model
selection. Having illustrated how to obtain a good detection function,
we show how to estimate abundance using that model, including
stratification. The final two sections of the article look at extensions
(both in terms of methodology and software) and put the package in a
broader context amongst other software packages used for estimating the
abundance of biological populations.

\section{Data setup}\label{data-setup}

The two example data sets used here are distributed with \pkg{Distance}
so readers can reproduce our results. However, data will be collected in
the field and formatted correctly for use with \pkg{Distance}. The
package allows for a flexible format for data input ranging from very
simple to complex:

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  In the simplest case, where the objective is to estimate a detection
  function, all that is needed is a vector of distances.
\item
  To include additional covariates into the detection function (see
  ``Detection functions'') we use a \code{data.frame}. The data.frame
  contains a column called \code{distance} and additional named columns
  for covariates potentially useful in modelling detectability (for
  example \code{observer} or \code{seastate}). Some column names are
  reserved: \code{object} for an observation identifier (see
  ``Extensions''), \code{size} for group or cluster size (see
  ``Detection functions'' and ``Abundance and variance estimation''),
  \code{detected} for whether an observation was detected (see
  ``Extensions'') and the columns described in the next bullet.
\item
  To estimate abundance beyond area covered by expended survey effort,
  additional information is required. This consists of information about
  which transect and occasion the observation was made (the
  \code{Sample.Label}), a column named \code{Effort} which gives the
  effort associated with that sample (for lines their length and for
  points the number of times that point was visited), the stratum the
  sample was located in (this may have any name and may be from pre- or
  post-survey stratification, see ``Estimating abundance and variance'')
  and strata areas (which has the same name as the stratum column,
  appended with \code{.Area}). We refer to this data format as
  ``flatfile'' where all information is contained in one table.
\end{itemize}

As we will see in ``Extensions'', further information is also required
when using more complex models.

@@

\begin{CodeChunk}
\begin{CodeInput}
library(Distance)
head(minke)
\end{CodeInput}
\begin{CodeOutput}
  Region.Label  Area Sample.Label Effort distance
1        South 84734            1  86.75     0.10
2        South 84734            1  86.75     0.22
3        South 84734            1  86.75     0.16
4        South 84734            1  86.75     0.78
5        South 84734            1  86.75     0.21
6        South 84734            1  86.75     0.95
\end{CodeOutput}
\end{CodeChunk}

@@

We will explore below the effects of including effort data during
analysis.

@@

The detection function describes the relationship between observed
distances and probability of detection. The detection function itself
models the probability
\(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
and is usually denoted \(g(y; \boldsymbol{\theta})\) where \(y\) is
distance (either from a line or point) and \(\boldsymbol{\theta}\) is a
vector of parameters to be estimated. Our goal is to estimate an
\emph{average probability of detection} (\(p\), average in the sense of
an average over distances), so we must integrate out distance (\(y\))
from the detection function: \[
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
\] where \(\pi(y)\) describes the distribution of objects with respect
to the sampler and \(\pi(x)=1/w\) for line transects and
\(\pi(r)=\frac{2r}{w^2}\) for point transects, taking into account the
geometry of the sampler (Buckland et al. 2001, Chapter 3) (letting \(x\)
denote a perpendicular distance from a line and \(r\) denote radial
distance from a point).

It is crucial that the detection function accurately models
detectability at small distances. We are less worried by its behaviour
further away from 0. To ensure that the model is not overly influence by
distances far from zero, we truncate the distances beyond a given
distance \(w\) (known as the \emph{truncation distance}). Fitting to
distances at great distances from zero does not demonstrably improve the
precision of abundance estimates (Buckland et al. 2001, 103--8,
151--53).

Models for the detection function are expected to have the following
properties (Buckland et al. 2015, Chapter 5):

\begin{itemize}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \emph{Shoulder}: we expect observers to be able to see objects near
  them, not just those directly adjacent to them. For this reason, we
  expect the detection function to be horizontal near zero distance.
\item
  \emph{Non-increasing}: we do not think that observers should be more
  likely to see things distant from the transect than those near the
  transect. If this occurs, it usually indicates an issue with field
  procedure (that the distribution of objects with respect to the line,
  \(\pi(y)\) is not what we expect), so we do not want the detection
  function to model this.
\item
  \emph{Model robust}: models should be flexible enough to make many
  different shapes.
\item
  \emph{Pooling robust}: many factors can affect the probability of
  detection and it is not possible to measure all of these. We would
  like models to produce unbiased results without inclusion of these
  factors.
\item
  \emph{Estimator efficiency}: we would like models to have low
  variances, but only given they satisfy the other properties above
  (which, if they are satisfied, would give low bias).
\end{itemize}

@@

There is a wide literature on possible formulations for the detection
function (Buckland 1992; Eidous 2005; Becker and Quang 2009; Giammarino
and Quatto 2014; Miller and Thomas 2015; Becker and Christ 2015).
\code{Distance} includes the most popular of these models. Here we will
detail the most popular detection function approach: ``key function plus
adjustments'' (K+A).

@@

Key function plus adjustment terms (or adjustment series) models are
formulated by taking a ``key'' function and optionally adding
``adjustments'' to it to improve the fit (Buckland 1992). Mathematically
we formulate this as: \[
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
\] where \(k\) is the key function and \(\alpha_O\) is sum series of
functions (given in Table \ref{tab:keyadj}), described as an
\emph{adjustment of order \(O\)}. Subscripts on the parameter vector
indicate those parameters belonging to each part of the model (i.e.
\(\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})\)).

Available models for the key are as follows: \[
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
\] Possible modelling options for key and adjustments are given in Table
\ref{tab:keyadj} and illustrated in Figure \ref{fig:hnhr} and
\ref{fig:keyadj}. We select the number of adjustment terms (\(K\)) by
AIC (further details in ``Model checking and model selection'').

@@

\begin{table}
\caption{Modelling options for key plus adjustment series models for the detection function.}
\begin{tabular}{llll}
\hline
Key function   & Form   & Adjustment series & Form\\
\hline
 Uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
 Half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
 Hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
\hline
\end{tabular}
\label{tab:keyadj}
\end{table}

When adjustment terms are used it may be necessary to standardise the
results to ensure that \(g(0)=1\), so we can redefine the detection
function as: \[
g(y; \boldsymbol{\theta}) = \frac{k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right)}{k(0; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(0; \boldsymbol{\theta}_\text{adjust})\right)}.
\]

A disadvantage of K+A models is that we must resort to constrained
optimisation (via the \pkg{Rsolnp} package) to ensure that the resulting
detection function is monotonic non-increasing over the whole range.

We do not always include adjustments (except in the case of the uniform
key), in which case we refer to ``key only'' models (see the next
section and ``Model checking and model selection'').

@@

\caption{Possible shapes for the detection function when adjustments are included for half-normal and hazard-rate models.\label{fig:keyadj}}\label{fig:adjust-mix-comp}
\end{figure}
\end{CodeChunk}

>>>>

\author{
David L Miller\\Integrated Statistics and University of St Andrews \AND Eric Rexstad\\University of St Andrews \And Len Thomas\\University of St Andrews \AND Laura Marshall\\University of St Andrews \And Jeffrey L Laake\\Marine Mammal Laboratory
}
\title{Distance Sampling in \proglang{R}}
\Keywords{distance sampling, abundance estimation, line transect, point transect, detection function, Horvitz-Thompson, \proglang{R}, Distance}

\Abstract{
Estimating the abundance and spatial distribution of animal and plant
populations is essential for conservation and management. We introduce
the \proglang{R} package \pkg{Distance} that implements distance
sampling methods to estimate abundance. We describe how users can obtain
estimates of abundance (and density) using the package as well
documenting the links it provides with other more specialized
\proglang{R} packages. We also demonstrate how \pkg{Distance} provides a
migration pathway from previous software, thereby allowing us to deliver
cutting-edge methods to the users more quickly.
}

\Plainauthor{David L Miller, Eric Rexstad, Len Thomas, Laura Marshall, Jeffrey L Laake}
\Plaintitle{Distance Sampling in R}
\Plainkeywords{distance sampling, abundance estimation, line transect, point transecs, detection function, Horvitz-Thompson, R, Distance}

@@

\Address{
    David L Miller\\
  Integrated Statistics and University of St Andrews\\
  16 Sumner Street, Woods Hole, MA 02543, USA and Centre for Research into
  Ecological and Environmental Modelling, The Observatory, St Andrews,
  Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:dave@ninepointeightone.net}{\nolinkurl{dave@ninepointeightone.net}}\\
  URL: \url{http://converged.yt}\\~\\
      Eric Rexstad\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:Eric.Rexstad@st-andrews.ac.uk}{\nolinkurl{Eric.Rexstad@st-andrews.ac.uk}}\\

      Len Thomas\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:len.thomas@st-andrews.ac.uk}{\nolinkurl{len.thomas@st-andrews.ac.uk}}\\
  URL: \url{http://lenthomas.org}\\~\\
      Laura Marshall\\
  University of St Andrews\\
  Centre for Research into Ecological and Environmental Modelling, The
  Observatory, St Andrews, Fife KY16 9LZ, Scotland\\
  E-mail: \href{mailto:lhm@st-andrews.ac.uk}{\nolinkurl{lhm@st-andrews.ac.uk}}\\

      Jeffrey L Laake\\
  Marine Mammal Laboratory\\
  Alaska Fisheries Science Center National Oceanic and Atmospheric
  Administration 7600 Sand Point Way N.E., Seattle, WA 98115, USA\\
  E-mail: \href{mailto:Jeff.Laake@noaa.gov}{\nolinkurl{Jeff.Laake@noaa.gov}}\\

@@

\usepackage{amsmath} \usepackage{amssymb} \usepackage{longtable}
\usepackage{booktabs}
\usepackage[colorinlistoftodos, color=green!40, textwidth=60, textsize=small]{todonotes}

@@

Distance sampling
\citep{Buckland:2001vm, Buckland:2004ts, buckland2015distance}
encompasses a suite of methods used to estimate the density and/or
abundance of biological populations. Distance sampling can be thought of
as an extension of plot sampling. Plot sampling involves selecting a
number of plots (small areas) at random within the study area and
counting the objects of interest that are contained within each plot. By
selecting the plots at random we can assume that the density of objects
in the plots is representative of the study area as a whole. One of the
key assumptions of plot sampling is that all objects within each of the
plots are counted. Distance sampling relaxes this assumption in that
observers are no longer required to detect (either by eye, video/audio
recording etc) and count everything within selected plots. While plot
sampling techniques are adequate for static populations occurring at
high density they are inefficient for more sparsely distributed
populations. Distance sampling provides a more efficient solution in
such circumstances.

Conventional distance sampling assumes the observer is located either at
a point or moving along a line and will observe all objects that occur
at the point or on the line. The further away an object is from the
point or line (also known as the sampler or transect) the less likely it
is that the observer will see it. We can use the distances to each of
the detected objects from the line or point to build a model of the
probability of detection given distance from the sampler -- the
\emph{detection function}. The detection function can be used to infer
how many objects were missed and thereby produce estimates of density
and/or abundance.

The Windows program Distance (or ``DISTANCE''; for clarity henceforth
``Distance for Windows'') can be used to fit detection functions to
distance sampling data. It was first released (versions 1.0 - 3.0;
principally programmed by Jeff Laake while working at the National
Marine Mammal Laboratory) as a console-based application (this in turn
was based on earlier software TRANSECT, \citet{Burnham:1980wz} and
algorithms developed in \citet{Buckland:1992fa}), before the first
graphical interface (Distance for Windows 3.5) was released in November
1998. Since this time it has evolved to include various design and
analysis features \citep{Thomas:2010cf}. Distance for Windows versions 5
onwards have included \proglang{R} \citep{rcore} packages as the
analysis engines providing additional, more complex analysis options
than those offered by the original (FORTRAN) code.

As Distance for Windows becomes increasingly reliant on analyses
performed in \proglang{R} and many new methods are being developed, we
are encouraging the use of our \proglang{R} packages directly.
\proglang{R} provides a huge variety of functionality for data
exploration and reproducible research, much more than is possible in
Distance for Windows.

Until now those wishing to use our \proglang{R} packages for straight
forward distance sampling analyses would have had to negotiate the
package \pkg{mrds} \citep{mrds-pkg} designed for mark-recapture distance
sampling \citep{Burt:2014gu}, requiring a complex data structure to
perform analyses. \pkg{Distance} is a wrapper package around \pkg{mrds}
making it easier to get started with basic distance sampling analyses in
\proglang{R}. The most basic detection function estimation only requires
a numeric vector of distances. Here we demonstrate how to use
\pkg{Distance} to fit detection functions, perform model checking and
selection, and estimate abundance.

@@

The distribution of the observed distances is a product of detectability
\citep[sometimes referred to as ``perception bias'';][]{Marsh:1989ho}
and the distribution of the animals with respect to the line or point.
Our survey design allows us to assume a distribution for the animals
with respect to the sampler.

For line transect studies we assume that objects are uniformly
distributed with respect to the line (i.e., the number of animals
available for detection is the same at all distances). For point
transect surveys, area increases linearly with radial distance, implying
a triangular distribution with respect to the point. Figure
\ref{fig:detectiondist} shows how these distributions, when combined
with a detection function, give rise to the observed distribution of
recorded distances.

@@

{\centering \includegraphics{paper_files/figure-latex/detectdists-1}

\caption{Panels show an example detection function (left), the pdf of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row). The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling. In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary). \label{fig:detectiondist}}\label{fig:detectdists}
\end{figure}
\end{CodeChunk}

Figure \ref{fig:pointslines} shows simulated sampling of a population of
500 objects using line and point transects and their corresponding
histograms of observed detection distances. Note that for the purposes
of distance sampling an ``object'' may either refer to an individual in
a population or a cluster of individuals.

\begin{CodeChunk}
\begin{CodeOutput}
Warning: package 'nlme' was built under R version 3.2.4
\end{CodeOutput}
\begin{figure}

@@

\caption{Left side plots show an example of a survey of a (1x1) area containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances (again, lines top and points bottom).\label{fig:pointslines}}\label{fig:points-and-lines}
\end{figure}
\end{CodeChunk}

Good survey design is essential to produce reliable density and
abundance estimates from distance sampling data. Survey design is beyond
the scope of this article but we refer readers to \citet[Chapter
7]{Buckland:2001vm} and \citet[Chapter 2]{buckland2015distance} for
introductory information; \citet{Strindberg:2004vr} contains information
on automated survey design using geographical information systems (GIS);
\citet{Thomas:2007wz} gives an example of designing a distance sampling
survey in a complex region. We also note that Distance for Windows
includes a GIS survey design tool \citep{Thomas:2010cf}.

\pkg{Distance} provides a selection of candidate functions to describe
the probability of detection and estimates the associated parameters
using maximum likelihood estimation. The probability of detecting an
object may not only depend on how far it is from the observer but also
on other factors such as weather conditions, ground cover, cluster size
etc. The \pkg{Distance} package also allows the incorporation of such
covariates into the detection function allowing the detection function
scale parameter to vary based on these covariates.

Having estimated the detection function's parameters, one can then
integrate out distance from the function (as the detection function
describes the probability of detection \emph{given} distance) to get an
``average'' probability of detection, which can be used to correct the
observed counts. Summing the corrected counts gives an estimate of
abundance in the area covered by surveys, which can be multiplied up to
the total study area.

In addition to randomly placed samplers distance sampling relies on
three other main assumptions. Firstly, all objects directly on the
transect line or point (i.e.~those at zero distance) are detected (see
``Extensions'' for methods to deal with the situation when this is not
possible). Secondly, objects are stationary or detected prior to any
movement. Thirdly, distance to the object must be measured accurately,
or the observation allocated to the correct distance bin for grouped
data. Depending on the survey species some of these assumptions may be
more difficult to meet than others. Further information on field methods
to help meet these assumptions can be found in \citet{Buckland:2001vm}
and \citet{buckland2015distance}.

The rest of the paper has the following structure: we describe data
formatting for \pkg{Distance}; candidate detection function models are
described in terms of formulation and fitted examples in \proglang{R}.
We then show how to perform model checking, goodness of fit and model
selection. We go on to show how to estimate abundance, including
stratified estimates of abundance. The final two sections of the article
look at extensions (both in terms of methodology and software) and put
the package in a broader context amongst other \proglang{R} packages
used for estimating the abundance of biological populations from
distance sampling data.

\section{Data}\label{data}

We introduce two example analyses performed in \pkg{Distance}: one line
transect and one point transect. These data sets have been chosen as
they represent typical data seen in practice. The below example analyses
are not intended to serve as guidelines, but to demonstrate features of
the software. Practical advice on approaches to analysis is given in
\citet{Thomas:2010cf}.

@@

The line transect data have been simulated from models fitted to
Antarctic minke whale (\emph{Balaenoptera bonaerensis}) data. These data
were collected as part of the International Whaling Commission's
International Decade of Cetacean Research Southern Ocean Whale and
Ecosystem Research (IWC IDCR-SOWER) programme 1992-1993 austral summer
surveys. These data consist of 99 observations that are stratified based
on location (near or distant from ice edge) and effort data (transect
lengths). Further details on the survey are available in
\citet{Branch:2001ua} (data is simulated based on the design used for
``1992/93 Area III'' therein).

@@

The point transect data set consists of 1485 observations of Amakihi
(\emph{Hemignathus virens}; a Hawaiian songbird), collected at 41 points
between 1992 and 1995. The data include distances and two covariates
collected during the survey: observer (a three level factor), time after
sunrise (transformed to minutes (continuous) or hours (factor)
covariates). Data are analysed comprehensively in
\citet{Marques:2007ey}.

\subsection{Data setup}\label{data-setup}

Generally, data collected in the field will require some formatting
before use with \pkg{Distance}, though there are a range of possible
formats, dependent on the model specification and the output required:

\begin{itemize}
\tightlist
\item
  In the simplest case, where the objective is to estimate a detection
  function and exact distances are collected, all that is required is a
  numeric vector of distances.
\item
  To include additional covariates into the detection function (see
  ``Detection functions'') a \code{data.frame} is required. Each row of
  the \code{data.frame} contains the data on one observation. The
  \code{data.frame} must contain a column named \code{distance}
  (containing the observed distances) and additional named columns for
  any covariates that may affect detectability (for example
  \code{observer} or \code{seastate}). The column name \code{size} is
  reserved for the cluster sizes (sometimes referred to as group sizes)
  in this case each row represents an observation of a cluster rather
  than individual (see \citet{Buckland:2001vm}, Section 1.4.3 for more
  on defining clusters and ``Extensions'' below for one approach to
  dealing with uncertain cluster size). Additional reserved names
  include \code{object} and \code{detected}, these are not required for
  conventional distance sampling and should be avoided (see
  ``Extensions'' for an explanation of their use).
\item
  To estimate density or to estimate abundance beyond the sampled area,
  additional information is required. Additional columns should be
  included in the \code{data.frame} specifying: \code{Sample.Label}, the
  ID of the transect; \code{Effort}, transect effort (for lines their
  length and for points the number of times that point was visited);
  \code{Region.Label}, the stratum containing the transect (which may be
  from pre- or post-survey stratification, see ``Estimating abundance
  and variance''); \code{Area}, the area of the strata. Transects which
  were surveyed but have no observations must be included in the data
  set with \code{NA} for \code{distance} and any other covariates. We
  refer to this data format (where all information is contained in one
  table) as ``flatfile'' as it can be easily created in a single
  spreadsheet.
\end{itemize}

As we will see in ``Extensions'', further information is also required
for fitting more complex models.

It is also possible to use distances collected in intervals (``binned''
or ``grouped'' data, as opposed to ``exact'' distances) to model the
detection function. In this case the column \code{distance} is replaced
by two columns \code{distbegin} and \code{distend} referring to the
distance interval start and end cutpoints. More information on binned
data is included in \citet{Buckland:2001vm} sections 4.5 and 7.4.1.2.

The columns \code{distance}, \code{Area} and (in the case of line
transects) \code{Effort} have associated units (though these are not
explicitly included in a \pkg{Distance} analysis). We recommend that
data in these columns are converted to SI units before starting any
analysis to ensure that resulting abundance and density estimates have
sensible units. For example, if distances from a line transect survey
are recorded in metres, the \code{Effort} columns should contain line
lengths also in metres and the \code{Area} column gives the stratum
areas in square metres. This would lead to density estimates of animals
per square metre.

@@

\begin{CodeChunk}
\begin{CodeInput}
library("Distance")
head(minke)
\end{CodeInput}
\begin{CodeOutput}
  Region.Label  Area Sample.Label Effort distance
1        South 84734            1  86.75     0.10
2        South 84734            1  86.75     0.22
3        South 84734            1  86.75     0.16
4        South 84734            1  86.75     0.78
5        South 84734            1  86.75     0.21
6        South 84734            1  86.75     0.95
\end{CodeOutput}
\end{CodeChunk}

@@

We will explore the consequences of including effort and stratum data in
the analysis below.

@@

The detection function models the probability
\(\mathbb{P}(\text{object detected } \vert \text{ object at distance } y)\)
and is usually denoted \(g(y; \boldsymbol{\theta})\) where \(y\) is
distance (from a line or point) and \(\boldsymbol{\theta}\) is a vector
of parameters to be estimated. Our goal is to estimate an \emph{average
probability of detection} (\(p\), average in the sense of an average
over distance from \(0\) to truncation distance \(w\)), so we must
integrate out distance (\(y\)) from the detection function: \[
p = \int_0^w \pi(y) g(y; \boldsymbol{\theta}) dy
\] where \(\pi(y)\) describes the distribution of objects with respect
to the sampler; \(\pi(x)=1/w\) for line transects and
\(\pi(r)=\frac{2r}{w^2}\) for point transects, taking into account the
geometry of the sampler \citep[usually referred to as the
\emph{probability density function of distances};][Chapter
3]{Buckland:2001vm}. When considering a particular transect type we let
\(x\) denote a perpendicular distance from a line and \(r\) denote
radial distance from a point (rather than using \(y\)).

It is crucial that the detection function accurately models
detectability at small distances; we are less worried by its behaviour
further away from 0. To ensure that the model is not overly influenced
by distances far from zero, we truncate the distances beyond a given
distance \(w\) (this can be done during analysis or while collecting
data in the field).

Models for the detection function are expected to have the following
properties \citep[Chapter 5]{buckland2015distance}:

\begin{itemize}
\tightlist
\item
  \emph{Shoulder}: we expect observers to be able to see objects near
  them, not just those directly in front of them. For this reason, we
  expect the detection function to be flat near the line or point.
\item
  \emph{Non-increasing}: we do not think that observers should be more
  likely to see distant objects than those nearer the transect. If this
  occurs, it usually indicates an issue with survey design or field
  procedures (for example that the distribution of objects with respect
  to the line, \(\pi(y)\) is not what we expect), so we do not want the
  detection function to model this
  \citep{Marques:2010he, Marques:2012fy, Miller:2015hw}.
\item
  \emph{Model robust}: models should be flexible enough to fit many
  different shapes.
\item
  \emph{Pooling robust}: many factors can affect the probability of
  detection and it is not possible to measure all of these. We would
  like models to produce unbiased results without inclusion of these
  factors.
\item
  \emph{Estimator efficiency}: we would like models to have low
  variances, given they satisfy the other properties above (which, if
  satisfied, would give low bias).
\end{itemize}

@@

There is a wide literature on possible formulations for the detection
function
\citep{Buckland:1992fa, Eidous:2005bj, Becker:2009cj, Giammarino:2014eg, Miller:2015hw, Becker:2015fi}.
\code{Distance} includes the most popular of these models. Here we
detail the most popular detection function approach: ``key function plus
adjustments'' (K+A).

@@

Key function plus adjustment terms (or adjustment series) models are
formulated by taking a ``key'' function and optionally adding
``adjustments'' to it to improve the fit \citep{Buckland:1992fa}.
Mathematically we formulate this as: \[
g(y; \boldsymbol{\theta}) = k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right),
\] where \(k\) is the key function and \(\alpha_O\) is sum series of
functions (given in Table \ref{tab:keyadj}), described as an
\emph{adjustment of order \(O\)}. Subscripts on the parameter vector
indicate those parameters belonging to each part of the model (i.e.
\(\boldsymbol{\theta} = (\boldsymbol{\theta}_\text{key}, \boldsymbol{\theta}_\text{adjust})\)).

Available models for the key are as follows: \[
k(y) = \left\{
\begin{array}{l l}
  \exp\left(-\frac{y^2}{2 \sigma^2}\right) & \quad \text{half-normal,} \\
  1-\exp\left(\left(-\frac{y}{\sigma}\right)^{-b}\right) & \quad \text{hazard-rate,} \\
  1/w & \quad \text{uniform.}
\end{array} \right.
\] Possible modelling options for key and adjustments are given in Table
\ref{tab:keyadj} and illustrated in Figures \ref{fig:hnhr} and
\ref{fig:keyadj}. We select the number of adjustment terms (\(K\)) by
AIC (further details in ``Model checking and model selection'').

@@

\begin{table}
\begin{tabular}{llll}
\hline
Key function   & Form   & Adjustment series & Form\\
\hline
 Uniform  & $1/w$   & cosine  & $\sum_{o=1}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=1}^O a_o (y/w)^{2o}$ \\
 Half-normal  & $\exp\left(-\frac{y^2}{2 \sigma^2}\right)$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Hermite polynomial & $\sum_{o=2}^O a_o H_{2o}(y/\sigma)$ \\
 Hazard-rate  & $1-\exp\left[-\left(\frac{y}{\sigma}\right)^{-b}\right]$ & cosine  & $\sum_{o=2}^O a_o \cos(o \pi y/w)$ \\
 & & Simple polynomial & $\sum_{o=2}^O a_o (y/w)^{2o}$ \\
\hline
\end{tabular}
\label{tab:keyadj}
\caption{Modelling options for key plus adjustment series models for the detection function. For each key function the default adjustments are cosine in \pkg{Distance}. Note that in the adjustments functions distance is divided by the width or the scale parameter to ensure the shape of adjustment independent of the units of $y$ (Marques et al., 2007); defaults are shown here, though either can be selected to rescale the distances.}
\end{table}

When adjustment terms are used it is necessary to standardise the
results to ensure that \(g(0)=1\), so we can redefine the detection
function as: \[
g(y; \boldsymbol{\theta}) = \frac{k(y; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(y; \boldsymbol{\theta}_\text{adjust})\right)}{k(0; \boldsymbol{\theta}_\text{key})\left( 1+ \alpha_O(0; \boldsymbol{\theta}_\text{adjust})\right)}.
\]

A disadvantage of K+A models is that we must resort to constrained
optimisation \citep[via the \pkg{Rsolnp} package;][]{rsolnp-pkg} to
ensure that the resulting detection function is monotonic non-increasing
over its range.

It is not always necessary to include adjustments (except in the case of
the uniform key) and in such cases we refer to these as ``key only''
models (see the next section and ``Model checking and model
selection'').

@@

\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.\label{fig:keyadj}}\label{fig:adjust-mix-comp}
\end{figure}
\end{CodeChunk}
